SHARP II: Dark Matter Substructure, Gravitational Lens flux-ratio Anomalies, and the Case of B1555+375: Beware the Effect of Edge-on Disks

### 2015-05-02 papers/B1555_EdgeOn/B1555_letter.tex
move some files

%While future large-scale surveys, such as the LSST, are expected to
%deliver new large samples of gravitationally lensed quasars, in the
%near future the discovery of new systems using narrow-line quasar
%emission \citep{N14} will provide new invaluable statistical
%perspective on current constraints based on flux-ratio anomalies.  
In this letter, we explore the idea that not all flux-ratio anomalies are
related to the presence of substructure, but that some originate in a
more complex lens galaxy mass distribution than initially
considered. To this end we use new infra-red data from the Strong
lensing at High Angular Resolution Program 
\citep[SHARP;][Fassnacht et al., in prep]{SHARP12}.
The main goal of the SHARP project is to re-observe known
quadruple and Einstein ring lens systems at high angular
resolution by making use of observations with (1) adaptive optics (AO)
on the Keck telescopes, (2) the Hubble Space Telescope, and (3) large
interferometric arrays \citep{SHARP12,V12}. In particular, to better
understand the causes of flux-ratio anomalies in gravitational lenses
we focus on the B1555+375 lens system\citep{Marlow99}.  This system was
discovered in the Cosmic Lens All-Sky Survey \citep{CLASS1,CLASS2} and
has a pair of merging images that show a strong flux-ratio anomaly at
radio wavelengths.  Follow-up observations ruled out radio propagation
effects as the cause of the anomaly \citep{K03,KD04}.  The B1555+375
system was included in both the observational and theoretical analysis
of \citet{Dalal2002} and \citet{Xu14} and, thus, understanding its
mass distribution is of key importance.  Our analysis makes use of
the new information obtained from the latest SHARP infrared imaging.
The paper is organized as follows: the  SHARP data are
introduced in section 2; in section 3 we present the lens model while
the results of this analysis and their implications are discussed in
section 4.

@@

As a first trial of modelling, we create a single SIE model to check
the performance of a simple lensing potential. Consistent with
previous studies \citep{Marlow99, Xu14}, our single elliptical
potential model cannot \textbf{reproduce} the flux anomaly in B1555+375.  We then
create two separate models of the lensing potential that both
incorporate the disk.  In the first we treat the lens as a SIE bulge\textbf{/halo}
and an exponential edge-on disk (SIE+expdisk).  In the second model
the disk is represented as a very elongated SIE component (2-SIE).
The lens system is red, with a strong jump in flux between the F814W
(roughly I) and F160W (roughly H) bands (Fig.~\ref{fig:multiband}), so
we assign a lens redshift of $z_\ell = 1.0$ and a source redshift of
$z_s =1.5$. \textbf{This assumption on redshifts is due to that some lens model parameters contain distance information. Our choice of redshifts does not affects the predicted magnifications nor flux-ratios.}

Our steps to explore the parameter space are: (1) Apply
strong priors from the SHARP AO image to the edge-on disk position angle
and ellipticity, and use the best-fit model of \citet{Marlow99} as the
starting point for the bulge\textbf{/halo} component and the source position. (2)
Let all parameters without priors vary to reach a temporary best
fit. (3) Start with the best fit in step 2 and let all parameters with
priors vary to reach another best fit (also ensure this best fit is
not far from the priors). (4) Let all parameters vary to reach the
final optimal model.  The full best-fit parameters are shown in Table
2. Both of these models produce lensed images with positions and
fluxes that are good matches to the radio observations.  While both of
the two component models produce significantly improved matches to the
fluxes than the single SIE model, the 2-SIE version has a somewhat
better performance than the SIE+expdisk model in reproducing flux
ratios. The comparison of the observations and our lens modelling is
given in Figure 3 and Tables 3 \& 4.  We have not included an external
shear component into our model because the constraints provided by the
image positions and fluxes would not be sufficient to constrain such a
model. However, we note that the image not affected by dust (image C)
in the Keck AO data is consistent with negligible shear (i.e. shear
strength $\Gamma_s=0.002$), when modeled with the pixelized technique
of \citet{V09}.

@@

\begin{table}
% \centering
% \begin{minipage}{140mm}
  \caption{Best-fit parameters of B1555+375 lens models. Subscripts 1 and 2 represent bulge\textbf{/halo} and disk components in each model respectively. Positions are offsets from radio image component A measured in units of arcsec. The velocity dispersion $\sigma$ is in units of $km/s$ and disk mass $M_{tot}$ is in units of $h^{-1} M_{\odot}$. $e$ is ellipticity and $\theta$ is the position angle measured east of north.}
  \begin{tabular}{@{}ccc}
\hline 
 Parameter  & \multicolumn{2}{c}{Lens model} \\
		&SIE+expdisk& 2-SIE		   
\\
\hline
$x_1$  	  & $-0.1794$	& $-0.1884$  \\
$y_1$	  &$-0.1881$	&$-0.1607$  \\

@@

From a theoretical perspective, \citet{Xu14} also find that substructures
cannot explain the observed radio flux anomalies in all the lenses
that were considered. After adding subhaloes from numerical simulations
to these systems, only one out of the eight radio-loud systems
achieves the level of the flux anomalies seen in the observations.  The rest of
them are unlikely (only $1 - 4$ per cent probability) to have flux
anomalies produced only by substructure, including our target B1555+375.
In fact, among those eight radio lenses, B1555+375 is one of the most
anomalous systems, but it is also one of the most challenging systems
to explain by substructures. \citet{Xu14} have concluded that the failure in
predicting flux-ratios is most likely due to oversimplified/improper lens
modelling. They also point out that the SIE component and the external
shear in their model are orthogonal, indicating a possible missing
ingredient in the lens model. Their results are consistent with the SHARP
$K^{\prime}$ band observations, in which an edge-on disk crosses the
radio merging double and which fully accounts for the observed flux-ratio anomaly.  In general, it is well known 
\citep[e.g,][]{Ka91} that lens models based only on the locations of lensed
quasar/AGN images have large degeneracies. Therefore, before
concluding that the flux-ratios in any given lens system are a result
of substructure, it is critical to explore a range of smooth models
that are informed by additional observations or to do a comprehensive
analysis such as that conducted by \citet{Xu14}. Otherwise, the derived
substructure fractions should be considered as upper limits.

@@

Our results clearly
demonstrate that the lack of knowledge on the lens galaxy mass
distribution, combined with the limited information provided by the
position of a few lensed images, can lead to improper lens models and
a misinterpretation of the origin of flux-ratio anomalies. \textbf{While future large-scale surveys, such as the LSST, are expected to
deliver new large samples of gravitationally lensed quasars, in the
near future the discovery of new systems using narrow-line quasar
emission \citep{N14} will provide new invaluable statistical
perspective on current constraints based on flux-ratio anomalies.} We stress,
therefore, that one should be very careful when interpreting the
nature of flux-ratio anomalies, as prior assumptions on the lens mass
distribution are a significant source of bias. Our analysis shows that
in order to derive precise quantification of the substructure
population and any derived properties from strongly lensed flux
anomalous systems, it is crucial to have a careful estimation of
non-substructure effects. In fact, even in the presence of mass
substructures, the observed flux-ratio anomalies could be still
dominated by the complex structure of the lens galaxy. Finally,
additional near-infrared AO imaging from SHARP shows that B1555+375 is
not the only system having a significant disk component detection in
the lensing galaxy.  These systems will be investigated in future
work, but the images alone strongly suggest that other lens systems
have similar links between flux-ratio anomalies and disk components.

@@

\section*{Acknowledgments}
S.V. is grateful to Dandan Xu for useful discussion on the nature of flux-ratio anomalies in high resolution numerical simulations.
\bibliographystyle{mn2e}
\bibliography{reference.bib}

>>>>

%While future large-scale surveys, such as the LSST, are expected to
%deliver new large samples of gravitationally lensed quasars, in the
%near future the discovery of new systems using narrow-line quasar
%emission \citep{N14} will provide new invaluable statistical
%perspective on current constraints based on flux-ratio anomalies.  
In this letter, we explore the idea that not all flux-ratio anomalies are
related to the presence of substructure, but that some originate in a
more complex lens galaxy mass distribution than initially
considered. To this end we use new infra-red data from the Strong
lensing at High Angular Resolution Program 
\citep[SHARP;][Fassnacht et al., in prep]{SHARP12, V12}.
The main goal of the SHARP project is to re-observe known
quadruple and Einstein ring lens systems at high angular
resolution by making use of observations with (1) adaptive optics (AO)
on the Keck telescopes, (2) the Hubble Space Telescope, and (3) large
interferometric arrays \citep{SHARP12,V12}. In particular, 
we focus on the B1555+375 lens system\citep{Marlow99} to better
understand the causes of flux-ratio anomalies in gravitational lenses.  This system was
discovered in the Cosmic Lens All-Sky Survey \citep{CLASS1,CLASS2} and
has a pair of merging images that show a strong flux-ratio anomaly at
radio wavelengths.  Follow-up observations ruled out radio propagation
effects as the cause of the anomaly \citep{K03,KD04}.  The B1555+375
system was included in both the observational and theoretical analysis
of \citet{Dalal2002} and \citet{Xu14} and, thus, understanding its
mass distribution is of key importance.  
%Our analysis makes use of
%the new information obtained from the latest SHARP infrared imaging.
The paper is organized as follows: the  SHARP data are
introduced in section 2; in section 3 we present the lens model while
the results of this analysis and their implications are discussed in
section 4.

@@

As a first trial of modelling, we create a single SIE model to check
the performance of a simple lensing potential. Consistent with
previous studies \citep{Marlow99, Xu14}, our single elliptical
potential model cannot \textbf{reproduce} the flux anomaly in B1555+375.  We then
create two separate models of the lensing potential that both
incorporate the disk.  In the first we treat the lens as a SIE bulge\textbf{/halo}
and an exponential edge-on disk (SIE+expdisk).  In the second model
the disk is represented as a very elongated SIE component (2-SIE). [WHY MODEL BULGE W/ SIE? COMMENT ON THE REASON FOR PICKING MODELS--LEON]
The lens system is red, with a strong jump in flux between the F814W
(roughly I) and F160W (roughly H) bands (Fig.~\ref{fig:multiband}), so
we assign a lens redshift of $z_\ell = 1.0$ and a source redshift of
$z_s =1.5$. \textbf{This assumption on redshifts is due to that some lens model parameters contain distance information. Our choice of redshifts does not affects the predicted magnifications nor flux-ratios.}

Our steps to explore the parameter space are: (1) Apply
strong priors from the SHARP AO image to the edge-on disk position angle
and ellipticity, and use the best-fit model of \citet{Marlow99} as the
starting point for the bulge\textbf{/halo} component and the source position. (2)
Let all parameters without priors vary to reach a temporary best
fit. (3) Start with the best fit in step 2 and let all parameters with
priors vary to reach another best fit (also ensure this best fit is
not far from the priors). (4) Let all parameters vary to reach the
final optimal model.  The \textbf{final} best-fit parameters are shown in Table
2. Both of these models produce lensed images with positions and
fluxes that are good matches to the radio observations.  While both of
the two component models produce significantly improved matches to the
fluxes than the single SIE model, the 2-SIE version has a somewhat
better performance than the SIE+expdisk model in reproducing flux
ratios. The comparison of the observations and our lens modelling is
given in Figure 3 and Tables 3 \& 4.  We have not included an external
shear component into our model because the constraints provided by the
image positions and fluxes would not be sufficient to constrain such a
model. However, we note that the image not affected by dust (image C)
in the Keck AO data is consistent with negligible shear (i.e. shear
strength $\Gamma_s=0.002$), when modeled with the pixelized technique
of \citet{V09}.

@@

\begin{table}
% \centering
% \begin{minipage}{140mm}
  \caption{Best-fit parameters of B1555+375 lens models. Subscripts 1 and 2 represent bulge\textbf{/halo} and disk components in each model respectively. Positions are offsets from radio image component A measured in units of arcsec. The velocity dispersion $\sigma$ is in units of $km/s$ and disk mass $M_{tot}$ is in units of $h^{-1} M_{\odot}$. \textbf{$e$ is ellipticity defined as the axis ratio $q=1-e$ }and $\theta$ is the position angle measured east of north.}
  \begin{tabular}{@{}ccc}
\hline 
 Parameter  & \multicolumn{2}{c}{Lens model} \\
		&SIE+expdisk& 2-SIE		   
\\
\hline
$x_1$  	  & $-0.1794$	& $-0.1884$  \\
$y_1$	  &$-0.1881$	&$-0.1607$  \\

@@

From a theoretical perspective, \citet{Xu14} also find that substructures
cannot explain the observed radio flux anomalies in all the lenses
that were considered. After adding subhaloes from numerical simulations
to these systems, only one out of the eight radio-loud systems
achieves the level of the flux anomalies seen in the observations.  The rest of
them are unlikely (only $1 - 4$ per cent probability) to have \textbf{observed distribution of flux
anomalies} produced only by substructure, including our target B1555+375.
In fact, among those eight radio lenses, B1555+375 is one of the most
anomalous systems, but it is also one of the most challenging systems
to explain by substructures. \citet{Xu14} have concluded that the failure in
predicting flux-ratios is most likely due to oversimplified/improper lens
modelling. They also point out that the SIE component and the external
shear in their model are orthogonal, indicating a possible missing
ingredient in the lens model. Their results are consistent with the SHARP
$K^{\prime}$ band observations, in which an edge-on disk crosses the
radio merging double and which fully accounts for the observed flux-ratio anomaly.  In general, it is well known 
\citep[e.g,][]{Ka91} that lens models based only on the locations of lensed
quasar/AGN images have large degeneracies. Therefore, before
concluding that the flux-ratios in any given lens system are a result
of substructure, it is critical to explore a range of smooth models
that are informed by additional observations or to do a comprehensive
analysis such as that conducted by \citet{Xu14}. Otherwise, the derived
substructure fractions should be considered as upper limits.

@@

Our results clearly
demonstrate that the lack of knowledge on the lens galaxy mass
distribution, combined with the limited information provided by the
position of a few lensed images, can lead to improper lens models and
a misinterpretation of the origin of flux-ratio anomalies. \textbf{While future large-scale surveys, such as the DES, KiDS-VIKING, Euclid, and LSST, are expected to
deliver new large samples of gravitationally lensed quasars, in the
near future the discovery of new systems using narrow-line quasar
emission \citep{N14} will provide new invaluable statistical
perspective on current constraints based on flux-ratio anomalies.} We stress,
therefore, that one should be very careful when interpreting the
nature of flux-ratio anomalies, as prior assumptions on the lens mass
distribution are a significant source of bias. Our analysis shows that
in order to derive precise quantification of the substructure
population and any derived properties from strongly lensed flux
anomalous systems, it is crucial to have a careful estimation of
non-substructure effects. In fact, even in the presence of mass
substructures, the observed flux-ratio anomalies could be still
dominated by the complex structure of the lens galaxy. Finally,
additional near-infrared AO imaging from SHARP shows that B1555+375 is
not the only system having a significant disk component detection in
the lensing galaxy.  These systems will be investigated in future
work, but the images alone strongly suggest that other lens systems
have similar links between flux-ratio anomalies and disk components.

@@

\section*{Acknowledgments}
S.V. is grateful to Dandan Xu for useful discussion on the nature of flux-ratio anomalies in high resolution numerical simulations.
L.V.E.K. is supported in part through an NWO-VICI career grant (project number 639.043.308).
\bibliographystyle{mn2e}
\bibliography{reference.bib}
